---
title: "lawCrimePlots.Rmd"
author: "Will W"
date: "11/8/2017"
output: html_document
---
### Data Preparation
> See "lawStaff_vs_Crime_Plot" R script

```{r, include=FALSE}
lawCrime <- read.csv("lawStaff_vs_Crime.csv", stringsAsFactors = FALSE)

names(lawCrime)
### Column Names ###
# "year"        "department"  "county"      "ft_so_m"     "ft_so_f"     "ft_c_m"     
# "ft_c_f"      "pt_so"       "pt_c"        "us_so_ft"    "us_so_pt"    "us_c_ft"    
# "us_c_pt"     "pop_covered" "so_per_1000" "num_res_so"  "total_crime"

# Replace null values with 0, from inconsistent data cleaning
columnsToReplace <- lawCrime[,c("us_so_ft", "us_so_pt", "num_res_so",
                                "us_c_ft", "us_c_pt")]
columnsToReplace[is.na(columnsToReplace)] <- 0
lawCrime[,c("us_so_ft", "us_so_pt", "num_res_so",
            "us_c_ft", "us_c_pt")] <- columnsToReplace

# Checking for rows with nulls,
# if nothings prints the data has no nulls
nulls <- complete.cases(lawCrime)
nulls
counter <- 0
for (null in nulls) {
  if (!null){
    print(paste("You have nulls at row", counter))
  }
  counter <- counter+1
}

# Create a total full time officer column,
# combining male and female records
lawCrime$total_so <- (lawCrime$ft_so_m + lawCrime$ft_so_f)
# Same for full time civilians
lawCrime$total_c <- lawCrime$ft_c_m + lawCrime$ft_c_f

################
### Map Prep ###
################

# Mapping county zip codes to names
postalCodes <- read.csv('us_postal_codes.csv')
colnames(postalCodes)[5] <- "county"
names(postalCodes)

# Just keep zip codes and county names, and states
postalCodes <- postalCodes[,c(-2,-4,-6,-7,-8)]
names(postalCodes)

# All county names match - whoo hoo!
lawCrime$county %in% postalCodes$county[postalCodes$State == 'Nebraska']
# Rows in dataset matching counties in lawCrime and located in NE
indexes <- which(postalCodes$State == 'Nebraska' & postalCodes$county %in% lawCrime$county) #& ])
# Keep only matching rows in the zipcode file
postalCodes <- postalCodes[indexes,]
# Keep only one zip for each County
postalCodes <- postalCodes[!duplicated(postalCodes$county),]
nrow(postalCodes)
# Double check each county still has a match
lawCrime$county %in% postalCodes$county[postalCodes$State == 'Nebraska']

# Create a new dataframe merging LawCrime with county zip codes
lawCrimeMapping <- merge(lawCrime, postalCodes, by = "county", type = "inner")
# Remove state and department columns
colnames(lawCrimeMapping)
lawCrimeMapping <- lawCrimeMapping[,-c(21, 3)]


#!!!!! Map only 2016 data !!!!!#
lawCrimeMapping <- lawCrimeMapping[lawCrimeMapping$year == 2016,]
# Aggregate a number of columns in the event you want to map different things
lawCrimeMapping <- aggregate(lawCrimeMapping[,c(3,4,5,6,7,8,9,10,12,13,14,17,18)], 
                             by=list(county=lawCrimeMapping$county), 
                             FUN=sum)
```

### Mapping the Data
> See "lawStaff_vs_Crime_Plot" R script

```{r, include=FALSE}
# Calling mapping libraries
library(readxl)
library(dplyr)
library(choroplethr)
library(choroplethrMaps)
library(ggplot2)
library(maps)
# Assigning a "region" to each county, 
# "region" is a ggplot specific mapping address
data(county.map)
# Keep Nebraska data only
county.map <- filter(county.map, STATE == 31)
# Keep only one region code per county name
county.map <- county.map[!duplicated(county.map$NAME),]
# Join by county name
lawCrimeMapping <- left_join(lawCrimeMapping, county.map, by = c("county" = "NAME"))
# value is a required mapping column name
colnames(lawCrimeMapping)[13] <- "value"
# Plotting by number of full time officers
county_mapSo <- county_choropleth(lawCrimeMapping, legend = "Number of Sworn Officers, 2016", 
                                state_zoom = "nebraska", num_colors=6)
# We can see there are many counties not officially covered by any police department
```

### Looking at Number of Sworn Officers by County
> Pretty interesting to see there are many counties which have no police department, our policing data does not reflect how resources are spread out for each one amongst neighboring counties.  Generally, there is one police department per county with 1 exception.  In the middle of the state, we see that the cluster of departments are responsible for patrolling the entire North to South region of the state.  

```{r, echo = FALSE}
county_mapSo
```


```{r, include=FALSE}
# Plotting by population covered
popCovered <- lawCrimeMapping[,c("region", "pop_covered")]
colnames(popCovered)[2] <- "value"
county_mapPop <- county_choropleth(popCovered, legend = "Pop Covered by County Departments, 2016", 
                                state_zoom = "nebraska", num_colors=6)
# We can only imagine how departments share coverage of counties they are not based in
```

### Population covered by County
> While we mapped by county, each county represents a department.  Here we can see the general population densities of Nebraska, as well as the additional burden placed on those central agencies - as they have a relatively high amount of population to cover given the size of the geographic area they operate in.

```{r}
county_mapPop
```

```{r, include=FALSE}
# Plotting by SO per 1000
SoPer1000 <- lawCrimeMapping[,c("region", "so_per_1000")]
colnames(SoPer1000)[2] <- "value"
county_SoPer1000 <- county_choropleth(SoPer1000, legend = "Number of Sworn Officers Per 1000, 2016", 
                                   state_zoom = "nebraska", num_colors=6)
# It makes sense the central part of the state, being the most spread out,
# has the lowest ratio
```

### SO per 1000 People by County
> This map does not translate well, as obviously departments, while not located in many counties have to go and investigate crimes in counties without a dedicated police department.  It is interesting there appears to be lower police presence on the panhandle portion of the state, while the central region enjoys the best officer to civilian ratio.

```{r, echo=FALSE}
county_SoPer1000
```

```{r, include=FALSE}
# Create a crime per population covered ratio column
lawCrime$crime_per_pop_covered <- lawCrime$total_crime / lawCrime$pop_covered

# Just looking at 2016
lawCrime2016 <- lawCrime[lawCrime$year == 2016,]
options(scipen=999)
# Evidently there are rural communities with crime rates,
# similiar to that of Omaha and Lincoln
plot(lawCrime2016$pop_covered, lawCrime2016$crime_per_pop_covered,
     main = "Comparing Department Population Covered to Crime Rate",
     xlab = "Population Covered", ylab = "Crime Rate Per 1000",
     xlim = c(0,500000))
colnames(lawCrime2016)
# Hiring trends by officer gender
soHires <- aggregate(lawCrime[,c(4,5)],
                   by=list(year=lawCrime$year),
                   FUN=sum)
```

### Annual SO Hiring Trends by Gender
> Hiring trends appear consistent statewide by gender, although only 5-10 female officers are being hired or leaving each year. Due to the relative involatile nature of this visualization, I'm not sure it's meaningful in our research. 

```{r, echo=FALSE}
plot(soHires$year, 
     soHires$ft_so_m,
     main = "Male SO Staffing Trends",
     xlab = "Year",
     ylab = "# NE Male Officers")

plot(soHires$year, 
     soHires$ft_so_f,
     main = "Female SO Staffing Trends",
     xlab = "Year",
     ylab = "# NE Female Officers")

```