---
title: "Visualization for County crime against Population and Staffing"
author: "Shivani, William, Chaitra"
date: "11/6/2017"
output: html_document
---
## Shivani - Scatter plot for Crime in County's of Nebraska against Population of the County

### Load the Required Libraries
```{r}
getwd()
```

```{r}
setwd("/Users/WillieWetz/Documents/GitHub/Data-Rangers/Deliverables/R_Plot_Deliverable")
library(ggplot2)
library(ggrepel)
```

### Load the dataset

```{r}
county_crime <- read.csv('County_crime_and_population_combined.csv')
lawCrime <- read.csv('lawStaff_vs_Crime.csv')
```


### Create a subset for counties with comparatively low population

```{r}
county_crime_with_low_population <- county_crime[county_crime$Population <= 10000 & county_crime$Population > 100,]
```


### Plot the scatter plot for the county's in the above subset

```{r}
p <- ggplot(county_crime_with_low_population, aes(Population, Total, color = County))

p + geom_point(shape=21, fill="White", size=3,stroke=1.5) + theme(legend.position = "none") + geom_text_repel(aes(label = County)) + xlab("Population of the County") + ylab("Total Crime from 2007 to 2016") + ggtitle("Scatterplot for Total crimes in the Nebraska County against Population")
```

### Final Plot with proper title and axis label

```{r}
p + geom_point(shape=21, fill="White", size=3,stroke=1.5) + theme(legend.position = "none") + geom_text_repel(aes(label = County)) + xlab("Population of the County") + ylab("Total Crime from 2007 to 2016") + ggtitle("Scatterplot for Total crimes in the Nebraska County against Population") + theme(plot.title = element_text(hjust = 0.5)) + scale_x_continuous(limits = c(100, 10000))
```

> Plot for Crime against population of the counties (with population less than 50000) shows many variation. As some of the counties like "Phelps" for an instance represents general idea of higher population and higher crime but then there are counties like Cedar and Knox which show the exact opposite with higher population and very less crime. This provides us an insight to explore other factors than population which might be a major factor on number of crimes in the counties.

> Using a smaller set of the data set gives more specific details from the visualization perspective and hence it was better to group a set of Counties and visualize crime data for that particular group. 

## Will - Mapping Law Enforcement Coverage

### Data Preparation
> See "lawStaff_vs_Crime_Plot" R script
### Set working Directory

```{r}
lawCrime <- read.csv('lawStaff_vs_Crime.csv')

# Replace null values with 0, from inconsistent data cleaning
columnsToReplace <- lawCrime[,c("us_so_ft", "us_so_pt", "num_res_so",
                                "us_c_ft", "us_c_pt")]
columnsToReplace[is.na(columnsToReplace)] <- 0
lawCrime[,c("us_so_ft", "us_so_pt", "num_res_so",
            "us_c_ft", "us_c_pt")] <- columnsToReplace

# Create a total full time officer column,
# combining male and female records
lawCrime$total_so <- (lawCrime$ft_so_m + lawCrime$ft_so_f)
# Same for full time civilians
lawCrime$total_c <- lawCrime$ft_c_m + lawCrime$ft_c_f

################
### Map Prep ###
################

# Mapping county zip codes to names
postalCodes <- read.csv('us_postal_codes.csv')
colnames(postalCodes)[5] <- "county"

# Just keep zip codes and county names, and states
postalCodes <- postalCodes[,c(-2,-4,-6,-7,-8)]

# Rows in dataset matching counties in lawCrime and located in NE
indexes <- which(postalCodes$State == 'Nebraska' & postalCodes$county %in% lawCrime$county) #& ])
# Keep only matching rows in the zipcode file
postalCodes <- postalCodes[indexes,]
# Keep only one zip for each County
postalCodes <- postalCodes[!duplicated(postalCodes$county),]


# Create a new dataframe merging LawCrime with county zip codes
lawCrimeMapping <- merge(lawCrime, postalCodes, by = "county", type = "inner")

# Remove state and department columns
lawCrimeMapping <- lawCrimeMapping[,-c(21, 3)]


#!!!!! Map only 2016 data !!!!!#
lawCrimeMapping <- lawCrimeMapping[lawCrimeMapping$year == 2016,]
# Aggregate a number of columns in the event you want to map different things
lawCrimeMapping$county <- as.character(lawCrimeMapping$county)

lawCrimeMapping$pop_covered <- as.numeric(lawCrimeMapping$pop_covered)

lawCrimeMapping <- aggregate(lawCrimeMapping[,c(3,4,5,6,7,8,9,10,12,13,14,17,18)], 
                             by=list(county=lawCrimeMapping$county), 
                             FUN=sum)
```

### Mapping the Data
> See "lawStaff_vs_Crime_Plot" R script

```{r, include=FALSE}
# Calling mapping libraries
library(readxl)
library(dplyr)
library(choroplethr)
library(choroplethrMaps)
library(ggplot2)
library(maps)
# Assigning a "region" to each county, 
# "region" is a ggplot specific mapping address
data(county.map)
# Keep Nebraska data only
county.map <- filter(county.map, STATE == 31)
# Keep only one region code per county name
county.map <- county.map[!duplicated(county.map$NAME),]
# Join by county name
lawCrimeMapping <- left_join(lawCrimeMapping, county.map, by = c("county" = "NAME"))
# value is a required mapping column name
colnames(lawCrimeMapping)[13] <- "value"
# Plotting by number of full time officers
county_mapSo <- county_choropleth(lawCrimeMapping, legend = "Number of Sworn Officers, 2016", 
                                state_zoom = "nebraska", num_colors=6)
# We can see there are many counties not officially covered by any police department
```

### Looking at Number of Sworn Officers by County
> Pretty interesting to see there are many counties which have no police department, our policing data does not reflect how resources are spread out for each one amongst neighboring counties.  Generally, there is one police department per county with 1 exception.  In the middle of the state, we see that the cluster of departments are responsible for patrolling the entire North to South region of the state.  

```{r, echo = FALSE}
county_mapSo
```


```{r, include=FALSE}
# Plotting by population covered
popCovered <- lawCrimeMapping[,c("region", "pop_covered")]
colnames(popCovered)[2] <- "value"
county_mapPop <- county_choropleth(popCovered, legend = "Pop Covered by County Departments, 2016", 
                                state_zoom = "nebraska", num_colors=6)
# We can only imagine how departments share coverage of counties they are not based in
```

### Population covered by County
> While we mapped by county, each county represents a department.  Here we can see the general population densities of Nebraska, as well as the additional burden placed on those central agencies - as they have a relatively high amount of population to cover given the size of the geographic area they operate in.

```{r}
county_mapPop
```

```{r, include=FALSE}
# Plotting by SO per 1000
SoPer1000 <- lawCrimeMapping[,c("region", "so_per_1000")]
colnames(SoPer1000)[2] <- "value"
county_SoPer1000 <- county_choropleth(SoPer1000, legend = "Number of Sworn Officers Per 1000, 2016", 
                                   state_zoom = "nebraska", num_colors=6)
# It makes sense the central part of the state, being the most spread out,
# has the lowest ratio
```

### SO per 1000 People by County
> This map does not translate well, as obviously departments, while not located in many counties have to go and investigate crimes in counties without a dedicated police department.  It is interesting there appears to be lower police presence on the panhandle portion of the state, while the central region enjoys the best officer to civilian ratio.

```{r, echo=FALSE}
county_SoPer1000
```

```{r, include=FALSE}
# Create a crime per population covered ratio column
lawCrime$crime_per_pop_covered <- lawCrime$total_crime / lawCrime$pop_covered

# Just looking at 2016
lawCrime2016 <- lawCrime[lawCrime$year == 2016,]
options(scipen=999)
# Evidently there are rural communities with crime rates,
# similiar to that of Omaha and Lincoln
plot(lawCrime2016$pop_covered, lawCrime2016$crime_per_pop_covered,
     main = "Comparing Department Population Covered to Crime Rate",
     xlab = "Population Covered", ylab = "Crime Rate Per 1000",
     xlim = c(0,500000),
     ylim = c(0,500000))
colnames(lawCrime2016)


lawCrime$department <- as.character(lawCrime$department)
lawCrime$county <- as.character(lawCrime$county)
lawCrime$pop_covered <- as.numeric(lawCrime$pop_covered)
# Hiring trends by officer gender
soHires <- aggregate(lawCrime[,c(4,5)],
                   by=list(year=lawCrime$year),
                   FUN=sum)
```

### Annual SO Hiring Trends by Gender
> Hiring trends appear consistent statewide by gender, although only 5-10 female officers are being hired or leaving each year. Due to the relative involatile nature of this visualization, I'm not sure it's meaningful in our research. 

```{r, echo=FALSE}
plot(soHires$year, 
     soHires$ft_so_m,
     main = "Male SO Staffing Trends",
     xlab = "Year",
     ylab = "# NE Male Officers")

plot(soHires$year, 
     soHires$ft_so_f,
     main = "Female SO Staffing Trends",
     xlab = "Year",
     ylab = "# NE Female Officers")

```


## Chaitra - Economic Indexes vs. Crime Categories

```{r, include=FALSE}
setwd("/Users/WillieWetz/Desktop")
```

```{r}
d <- read.csv("test.csv")
```

```{r, include=FALSE}
pairs = d[, c("S_P_500_Percent_Gain_or_Loss", "Crime_Unknown", "Crime_Criminal_Homicide")]
# pairs(d, labels = colnames(d), main = "Pair matrix ", pch = 21, bg = c("red", "green3", "blue")[unclass(d$lith)], upper.panel = NULL)

## Performing Correlation and Plotting the data ##

### Relation between the S&P percent gain/loss and the homicide crimes committed
cor.test(d$S_P_500_Percent_Gain_or_Loss, d$Crime_Criminal_Homicide)
library(car)
```

```{r, echo=FALSE}
plot(d$S_P_500_Percent_Gain_or_Loss, d$Crime_Criminal_Homicide, type='h', main="Rel btwn S&P Percent and homicide crime",col="red", xlab="S&P Percent Gain/Loss", ylab="Criminal Homicide Crime", col.lab="blue", cex=2, cex.lab=0.85)
```

### Relation between the S&P percent gain/loss and the crimes caused due to negligence
```{r, echo=FALSE}
#cor.test(d$S_P_500_Percent_Gain_or_Loss, d$Crime_Death_by_Negligence)
plot(d$S_P_500_Percent_Gain_or_Loss, d$Crime_Death_by_Negligence, type='h', main="Rel btwn S&P Percent and crimes due to negligence",col="red", xlab="S&P Percent Gain/Loss", ylab="Crimes caused due to negligence", col.lab="blue", cex=2, cex.lab=0.85)
```

### Relation between the S&P percent gain/loss and the  rapes committed
```{r, echo=FALSE}
#cor.test(d$S_P_500_Percent_Gain_or_Loss, d$Crime_Forcible_Rape)
plot(d$S_P_500_Percent_Gain_or_Loss, d$Crime_Forcible_Rape, type='h', main="Rel btwn S&P Percent and forcible rapes",col="red", xlab="S&P Percent Gain/Loss", ylab="Forcible Rape Crimes", col.lab="blue", cex=2, cex.lab=0.85)
```

### Relation between the S&P percent gain/loss and the crimes resulting from robbery
```{r, echo=FALSE}
#cor.test(d$S_P_500_Percent_Gain_or_Loss, d$Crime_Robbery)
plot(d$S_P_500_Percent_Gain_or_Loss, d$Crime_Robbery, type='h', main="Rel btwn S&P Percent and crimes_robbery ",col="red", xlab="S&P Percent Gain/Loss", ylab="Crimes Robbery ", col.lab="blue", cex=2, cex.lab=0.85)
```

### Relation between the S&P percent gain/loss and the crimes resulting from assault
```{r, echo=FALSE}
#cor.test(d$S_P_500_Percent_Gain_or_Loss, d$Crime_Aggravated_Assault)
plot(d$S_P_500_Percent_Gain_or_Loss, d$Crime_Aggravated_Assault, type='h', main="Rel btwn S&P Percent and aggravated assault ",col="red", xlab="S&P Percent Gain/Loss", ylab="Aggravated Assault Crimes ", col.lab="blue", cex=2, cex.lab=0.85)
```

### Relation between the S&P percent gain/loss and the crimes resulting from burglary
```{r, echo=FALSE}
#cor.test(d$S_P_500_Percent_Gain_or_Loss, d$Crime_Burglary)
plot(d$S_P_500_Percent_Gain_or_Loss, d$Crime_Burglary, type='h', main="Rel btwn S&P Percent and burglary",col="red", xlab="S&P Percent Gain/Loss", ylab="Burglary Crimes ", col.lab="blue", cex=2, cex.lab=0.85)
```

### Relation between the S&P percent gain/loss and driving under the influence
```{r, echo=FALSE}
#cor.test(d$S_P_500_Percent_Gain_or_Loss, d$Crime_Driving_Under_the_Influence)
plot(d$S_P_500_Percent_Gain_or_Loss, d$Crime_Driving_Under_the_Influence, type='h', main="Rel btwn S&P Percent and driving under the influence ",col="red", xlab="S&P Percent Gain/Loss", ylab="Driving under the influnce Crimes", col.lab="blue", cex=3, cex.lab=0.90)
### p-value=0.8262, thus there exists a correlation between the S&P percent gain/loss and the crimes committed under the influence.
```

### Analysis Results ###

> From the above analysis, it is observed that the S&P Percent Gain or Loss impacts/influences the crimes caused the under the influence. 


### Data Description ###

> The unemployment data in Nebraska is collected from 2007-2016 along with the types of crimes committed in those paticular years. Here, we are analysing the relation between the unemployment rate and the the types of crimes committed. This analysis will help us identify patterns in our datasets.

### Loading the dataset ###
```{r, include=FALSE}
d <- read.csv("test.csv")
```

### Relation between the unemeployment rate in Nebraska and the crimes resulting from simple assault
```{r}
plot(d$NE_Unemployment_Rate, d$Crime_Simple_Assault, type='h', main="Rel btwn NE unemp_rate and simple assault",col="red", xlab="NE Unemployment Rate", ylab="Simple Assault Crimes ", col.lab="blue", cex=3, cex.lab=0.90)  
```

### Relation between the unemeployment rate in Nebraska and drug abuse related crimes
```{r}
plot(d$NE_Unemployment_Rate, d$Crime_Drug_Abuse_Violations, type='h', main="Rel btwn NE unemp_rate and drug abuse violation ",col="red", xlab="NE Unemployment Rate", ylab="Drug Abuse Violation Crimes", col.lab="blue", cex=3, cex.lab=0.90)
```

### Relation between the unemeployment rate in Nebraska and curfew related crimes
```{r}
plot(d$NE_Unemployment_Rate, d$Crime_Curfew, type='h', main="Rel btwn NE unemp_rate and curfew ",col="red", xlab="NE Unemployment Rate", ylab="Cufew related Crimes", col.lab="blue", cex=3, cex.lab=0.90)
```

### Analysis Results ###

> From the above analysis, it is observed that the unemployment rate in NE affects/influences the crimes resulting from assault, drug abuse related crimes and curfew related crimes. 


