---
title: "Unemployment rate in NE Vs types of crimes"
author: "Chaitra"
date: "November 7, 2017"
output:
  html_document: default
  pdf_document: default
---

######################################
### Created by Chaitra Venkatesan ###
### Released under the MIT license ###
### 8086 R Plot Deliverable ###
### 11-6-17###
### Impact of the unemployment rate in Nebarska on the types of crimes committed ####
######################################
### Preparation ###
###################

### Data Description ###

> The unemployment data in Nebraska is collected from 2007-2016 along with the types of crimes committed in those paticular years. Here, we are analysing the relation between the unemployment rate and the the types of crimes committed. This analysis will help us identify patterns in our datasets.

### Loading the dataset ###
```{r, include=FALSE}
d <- read.csv("test.csv")
```

### Relation between the unemeployment rate in Nebraska and the crimes resulting from simple assault
```{r}
plot(d$NE_Unemployment_Rate, d$Crime_Simple_Assault, type='h', main="Rel btwn NE unemp_rate and simple assault",col="red", xlab="NE Unemployment Rate", ylab="Simple Assault Crimes ", col.lab="blue", cex=3, cex.lab=0.90)  
```

### Relation between the unemeployment rate in Nebraska and drug abuse related crimes
```{r}
plot(d$NE_Unemployment_Rate, d$Crime_Drug_Abuse_Violations, type='h', main="Rel btwn NE unemp_rate and drug abuse violation ",col="red", xlab="NE Unemployment Rate", ylab="Drug Abuse Violation Crimes", col.lab="blue", cex=3, cex.lab=0.90)
```

### Relation between the unemeployment rate in Nebraska and curfew related crimes
```{r}
plot(d$NE_Unemployment_Rate, d$Crime_Curfew, type='h', main="Rel btwn NE unemp_rate and curfew ",col="red", xlab="NE Unemployment Rate", ylab="Cufew related Crimes", col.lab="blue", cex=3, cex.lab=0.90)
```

### Analysis Results ###

> From the above analysis, it is observed that the unemployment rate in NE affects/influences the crimes resulting from assault, drug abuse related crimes and curfew related crimes. 



